\section{ResearchArena Benchmark}

\ra\ evaluates whether off-the-shelf CLI agents can complete an end-to-end research pipeline under a minimal scaffold. The scaffold provides domain seeds, resource limits, stage prompts, and review instructions, but it does not provide a hand-designed research method, curated experimental codebase, or fixed paper template beyond normal conference-style expectations.

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
  node distance=7mm,
  box/.style={draw, rounded corners=2pt, align=center, minimum height=8mm, inner xsep=5pt, fill=gray!6},
  wide/.style={draw, rounded corners=2pt, align=center, minimum height=8mm, inner xsep=5pt, fill=blue!5},
  arrow/.style={-Latex, thick}
]
\node[box] (seed) {Domain\\seed};
\node[box, right=of seed] (agent) {CLI\\agent};
\node[box, right=of agent] (idea) {Ideation\\+ self-review};
\node[box, right=of idea] (exp) {Code, configs,\\logs, results};
\node[box, right=of exp] (paper) {Paper writing\\+ self-review};
\node[box, right=of paper] (final) {Final artifacts};
\node[wide, below=12mm of exp] (eval) {PDF-only \sar\quad+\quad Artifact-grounded \pr\quad+\quad Manual integrity audit};
\draw[arrow] (seed) -- (agent);
\draw[arrow] (agent) -- (idea);
\draw[arrow] (idea) -- (exp);
\draw[arrow] (exp) -- (paper);
\draw[arrow] (paper) -- (final);
\draw[arrow] (final.south) |- (eval.east);
\draw[arrow] (exp.south) -- (eval.north);
\end{tikzpicture}
}
\caption{\ra\ workflow. Agents proceed from a domain seed through ideation, experiments, paper writing, self-review, and final evaluation. Artifact-grounded review uses the paper together with code, logs, results, and references.}
\label{fig:workflow}
\end{figure}

\begin{table}[t]
\centering
\small
\begin{tabularx}{\linewidth}{l l c c c c}
\toprule
Agent & Model & CPU seeds & GPU seeds & Trials/seed & Papers \\
\midrule
Claude Code & Opus 4.6 & 5 & 8 & 3 & 39 \\
Codex & GPT-5.4 & 5 & 8 & 3 & 39 \\
Kimi Code & K2.5 & 5 & 8 & 3 & 39 \\
\midrule
Total & -- & 5 & 8 & 3 & 117 \\
\bottomrule
\end{tabularx}
\caption{Benchmark setup. Each agent generates one paper for each trial in each domain seed. CPU seeds are causal learning, compiler optimization, data integration and cleaning, operating-system design, and probabilistic methods. GPU seeds are AI for biology, computer vision, datasets and benchmarks, generative models, interpretability of learned representations, NLP, privacy in ML, and supervised representation learning.}
\label{tab:setup}
\end{table}

\paragraph{Agents and domains.}
Table~\ref{tab:setup} summarizes the study design. We evaluate Claude Code, Codex, and Kimi Code. Each agent produces 39 papers: three trials for each of 13 domain seeds. The CPU-only domains emphasize algorithmic, systems, and data-management tasks that can be run without accelerators. The GPU domains emphasize modern ML areas where training, fine-tuning, or model evaluation is common.

\paragraph{Hardware.}
Main experiments use either CPU-only resources or one NVIDIA RTX A6000 GPU with 48GB memory. Each agent receives 4 CPU cores and 60GB RAM. We additionally run a Codex GPU scaling experiment on NVIDIA H100 GPUs with 80GB memory while holding other settings fixed.

\paragraph{Pipeline.}
Each run follows the same high-level stages. During ideation, the agent proposes a research question and experiment plan, with up to three rounds of self-review and revision. During experimentation, the agent writes and executes code, records logs, and stores result artifacts, again with self-review. During paper writing, the agent drafts the manuscript and revises it through self-review. Finally, the paper is evaluated by \sar\ and by artifact-grounded peer review.

\paragraph{Collected artifacts.}
For each generated paper, \ra\ retains the final PDF or source manuscript, experiment code, configuration files, execution logs, result files such as JSON or CSV outputs, generated figures, self-reviews, peer reviews, Stanford Agentic Reviewer output, and metadata. The retained artifacts make it possible to evaluate not only what the paper says, but whether the evidence in the workspace supports it.
