\section{Case Studies}

Aggregate scores hide important differences in how generated papers fail. Table~\ref{tab:cases} summarizes six representative cases. The cases are not meant as isolated anecdotes; they illustrate recurring artifact-grounded failure modes that directly affect scientific validity.

\begin{table}[t]
\centering
\scriptsize
\begin{tabularx}{\linewidth}{p{0.11\linewidth} p{0.22\linewidth} p{0.18\linewidth} X}
\toprule
Agent & Paper or topic & Artifact-grounded issue & Lesson \\
\midrule
Claude Code & Entropy-guided adaptive token merging & Meaningful mechanistic insight but weak evidence & Strong framing and useful analysis do not compensate for limited baselines and narrow validation. \\
Claude Code & \emph{The Algebra of Compiler Passes} & Claimed 87 benchmarks, artifacts support 20 & Artifact inspection changes the scale and credibility of the measurement claim. \\
Codex & CLIP/CIFAR-10 style controlled study & Narrow, careful negative result & Lower fabrication risk can coexist with limited scope and modest significance. \\
Kimi Code & Hard-coded benchmark statistics and fabricated outputs & Reported benchmark evidence cannot be reconstructed from execution artifacts & Plausible tables are insufficient without provenance through logs and result files. \\
Kimi Code & CXL-aware CPU scheduling & Method-code mismatch & A named systems method may not be the method actually implemented or evaluated. \\
Claude Code & Shapley-style compiler pipeline analysis & Missing relevant baseline similar to ShapleyPipe & Even faithful artifacts can support an incomplete scientific comparison. \\
\bottomrule
\end{tabularx}
\caption{Representative case studies. Artifact-grounded review distinguishes weak evidence, scale overclaiming, narrow but faithful execution, fabricated outputs, method-code mismatch, and missing baselines.}
\label{tab:cases}
\end{table}

The most common pattern is not a total absence of work. In several cases, the agent implemented code, ran experiments, and wrote a coherent paper. The failure is that the final paper overstates what the artifacts support. For example, a benchmark paper can claim a broad sweep while the released pipeline covers a much smaller subset; a systems paper can describe a method whose code path omits a claimed component; and an ML paper can mix evaluation protocols while presenting them as a single comparison.

These cases explain why artifact-grounded review changes the evaluation target. A PDF-only reviewer may reasonably judge a paper by clarity, positioning, and apparent empirical results. An artifact-grounded reviewer can ask whether the claimed benchmark size appears in the code, whether the logs show the reported runs, whether result files contain the table entries, and whether references resolve to real publications. The same manuscript can therefore receive a plausible PDF-only score and still fail as a faithful account of the conducted research.
