\documentclass{article}

\usepackage[preprint]{neurips_2026}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{array}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage{url}
\usepackage{hyperref}
\usetikzlibrary{arrows.meta,positioning,fit}

\graphicspath{{figures/}}
\newcommand{\ra}{ResearchArena}
\newcommand{\sar}{SAR}
\newcommand{\pr}{PR}
\newcommand{\agent}[1]{\textsc{#1}}
\newcommand{\yes}{Yes}
\newcommand{\no}{No}
\newcommand{\maybe}{Partial}

\title{ResearchArena: Process-Grounded Evaluation of End-to-End Research by CLI Agents}

\author{Anonymous Authors}

\begin{document}

\maketitle

\begin{abstract}
Auto-research is not just paper generation. A research agent must form an idea, implement and run experiments, interpret outputs, cite prior work, and write a manuscript that remains faithful to the artifacts produced by the process. We introduce \ra, a benchmark and empirical study of off-the-shelf command-line agents performing an end-to-end research pipeline under a minimal scaffold. \ra\ contains 117 agent-generated papers from Claude Code, Codex, and Kimi Code across 13 computer-science domains, together with code, configurations, execution logs, result files, reviews, and metadata. We also introduce artifact-grounded peer review, an evaluation protocol in which reviewers inspect the final paper together with the code, logs, result artifacts, and references. This protocol evaluates paper-artifact faithfulness rather than only paper plausibility. Across the study, current CLI agents often complete research-like artifacts, but completion does not imply validity: experimental rigor, scientific significance, reference integrity, and paper-artifact faithfulness remain the dominant bottlenecks. PDF-only review assigns substantially higher apparent quality than artifact-grounded review, especially for papers whose reported settings or results cannot be reconstructed from the released artifacts.
\end{abstract}

\input{sections/intro}
\input{sections/related_work}
\input{sections/benchmark}
\input{sections/artifact_grounded_review}
\input{sections/results}
\input{sections/case_studies}
\input{sections/discussion}
\input{sections/limitations}
\input{sections/conclusion}

\bibliographystyle{plainnat}
\bibliography{references}

\clearpage
\appendix
\input{sections/appendix}

\end{document}
